miniKanren with fair search strategies

### 2019-05-05 Paper/paper.tex
intro

\newcommand{\conde}{\texttt{cond$^e$} }
\newcommand{\conj}{\texttt{conj}}
\newcommand{\disj}{\texttt{disj}}
\newcommand{\clisting}[1]{
\begin{center}
  \begin{tabular}{c}
	\lstinputlisting{#1}
    \end{tabular}
\end{center}
}
\newcommand{\DFSi}[0]{
  DFS$_\textrm{i}$
}

@@

The syntax of a programming language should reflect its semantics. When writing 
a \conde expression in miniKanren, a programmer would expect all clauses share 
the same chance of being explored, as these clauses are written in 
parallel. The existing search strategy, interleaving DFS (DFS$_\textrm{i}$), 
however, prioritize its clauses by the order how they are written down. 
Similarly, when a \conde is followed by another goal conjunctively, a 
programmer would expect answers in parallel share the same chance of being 
explored. Again, the answers by DFS$_\textrm{i}$ is different from the 
expectation. We have devised three new search strategies that have different 
level of fairness in \disj and \conj.

@@

\conde is the flagship disjunction operator in miniKanren. A subtlety arises
when a \conde contains many disjunction clauses: not every clause has the equal
chance to contribute to the result. As an example, consider the following
relation \texttt{repeato} and its invocation.
\clisting{Figures/repeato.rkt}
Next, consider the following disjuction of invoking \texttt{repeato} with
four different letters.
\clisting{Figures/example.rkt}
\conde intuitively corresponds to \texttt{or} in logic. And thus an unsuspicious
beginner would expect each letter to contribute equally to the result, as
following.
\clisting{Figures/run-repeato-fair.rkt}
The \conde in TR2, however, generates a less expected result.
\clisting{Figures/run-repeato-idfs.rkt}

\conde implements interleaving DFS(\DFSi), the cause of this unexpected result.
In a \conde, each clause takes a half of its received computational resources
and pass the other half to its following clauses, except for the last clause
that takes all resources it receives. In the example above, the letters
\texttt{c} and \texttt{d} are allocated with less resources than \texttt{a},
and thus \texttt{c} and \texttt{d} barely contribute to the result.

\DFSi is powerful for an expert. By carefully organizing the order of \conde
clauses, a miniKanren program can explore more ``interesting'' clauses than
those uninteresting ones, and thus use computational resources efficiently.
A little miniKanrener, however, may beg to differ--understanding
implementation details and fiddling with clauses order is not the first
priority of a beginner.

There is another reason that miniKanren could use more search strategies than
just \DFSi. In many applications, there does not exist one order that serves
for all purposes. For example, a relational dependent type checker contains
clauses for constructors that build data and clauses for eliminators that use
data. When this type checker is used to generate simple and shallow programs,
the clauses of constructors should be put in the front of \conde.
When performing proof search for complicated programs, the clauses of eliminator
should take the focus. With \DFSi, these two uses can not be efficient at the
same time. In fact, to make on use efficient, the other one must be drastically
slow.

\NOTE[LKC]{Should I explain ``state'' and ``search space'' here?}

Under the hood, \conde clauses are combined with \disj{}. The \disj{} 
implementation in \citep{friedman_reasoned_2018} is not \emph{fair}.

\begin{defn}[fair \disj]
	A \disj{} is fair iff it allocates computational resource evenly among 
	search spaces derived from disjunctive goals. \citet{seres1999algebra}
\end{defn}

A closely related concept is \emph{almost-fair} \disj.

\begin{defn}[almost-fair \disj]
	A \disj{} is almost-fair iff it allocates computational resource so evenly 
	among search spaces derived from disjunctive goals that the maximal ratio 
	of resources is bounded by a constant.
\end{defn}

The \disj{} in \citet{friedman_reasoned_2018}, or more precisely, in its search 
strategy, interleaving DFS(DFS$_\textrm{i}$), is neither fair nor almost-fair. 
Fair depth-first search (DFS$_\textrm{f}$), a new search strategy in this 
paper, has fair \disj. And balanced interleaving depth-first search 
(DFS$_\textrm{bi}$), another new strategy, has almost-fair \disj. Breath-first 
search (BFS), a known strategy \citet{seres1999algebra}, has fair \disj{} and 
\emph{fair \conj{}}.

\begin{defn}[fair \conj]
	A \conj{} is fair iff it allocates computational resource evenly among 
	search spaces derived from states in the same bag, where bags are finite 
	list of states.
\end{defn}

A comparison of fairness of search strategies is in Fig.~\ref{fairness}.

To summarize our contribution, we
\begin{itemize}
	\item propose a new concept, almost-fair \disj{}.
	\item propose a new definition of fair \conj{}.
	\item propose and implement balanced interleaving depth-first search 
	(DFS$_\textrm{bi}$), a new search strategy with almost-fair \disj{}.
	\item propose and implement fair depth-first search (DFS$_\textrm{f}$), a 
	new search strategy with fair \disj{}.
	\item implement in a new way breath-first search (BFS), a search strategy 
	with fair \disj{} and fair \conj{}
	(our code runs faster in all benchmarks and is simpler)
	\item prove our BFS implementation is equivalent with the one by 
	\citet{seres1999algebra}.
\end{itemize}

\begin{figure}[tbp]
	\begin{tabular}{|c|c|c|c|c|}
		\hline 
		fairness & DFS$_\textrm{i}$ & DFS$_\textrm{bi}$ & DFS$_\textrm{f}$ & BFS \\ 
		\hline 
		disj & unfair & almost-fair & fair & fair \\ 
		\hline 
		conj & unfair & unfair & unfair & fair \\ 
		\hline 
	\end{tabular} 
	\caption{fairness of search strategies}
	\label{fairness}
\end{figure}

% Interleaving DFS, the search strategy of \conde in
% miniKanren~\citet{friedman_reasoned_2018}, allocates computational resource
% by the order in which the clauses are written. Each clause of a
% \conde takes half of the current resource and passes the other half
% to its following clauses, except for the last clause that takes all of the 
%current resource. 
% The biased treatment provides both opportunity and burden: miniKanren users 
%can place more frequently used 
% goal at the beginning to optimize their programs; however, it might be a 
% catastrophe if a goal that generates many useless states is placed before 
%more 
% important goals. Seasoned miniKanreners usually know how to utilize 
% the unfairness to optimize their programs. However, we believe search 
%strategies that is less sensitive to goal order can also be useful to little 
%miniKanreners as well as seasoned ones. We propose two such search strategies, 
%balanced interleaving DFS (DFS$_\textrm{bi}$) and breadth-first search (BFS), and observe 
%how they affect the efficiency and the answer order of known miniKanren 
%programs. The experiment is conducted with the miniKanren from \textit{The 
%Reasoned Schemer, 2nd Edition}.

@@

In this section, we elaborate fairness by running queries about 
\texttt{repeato2}, a relational definition that relates a term \texttt{x} with a 
non-empty list whose elements are \texttt{x} (Fig.~\ref{repeato}).

\begin{figure}
	\lstinputlisting{Figures/repeato.rkt}
	\caption{\texttt{repeato} and an example run}
	\label{repeato2}
\end{figure}

@@

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the answer. DFS$_\textrm{i}$, the current search
strategy, however, results in many more lists of \texttt{a}s than lists
of other letters. And some letters, e.g. \texttt{c} and \texttt{d}, are
rarely seen. The situation would be exacerbated if \conde contains more clauses.

@@

We borrow the definition of \emph{fair} \disj{} from \citet{seres1999algebra}: 
search strategies with \emph{fair} \disj{} should allocate resources evenly 
among disjunctive goals. Running the same program with DFS$_\textrm{f}$ and BFS 
give the following result.

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-disj-fair.rkt}
	\end{tabular}
\end{center}

Now we are in a middle place between fair and unfair -- search strategies
with  \emph{almost-fair} \disj{} should allocate resources so evenly among
disjunctive goals that the maximal ratio of resources is bounded by a constant. 
Our new search strategy,
DFS$_\textrm{bi}$, has almost-fair \disj{}. It is
fair when the number of goals is a power of 2, otherwise, some goals are 
allocated twice as many resources than the others. In the previous example,
DFS$_\textrm{bi}$
gives the same result. And in the following example, where the \conde has 5 
clause, the clauses of \texttt{b}, \texttt{c}, and \texttt{d} are allocated 
more resources.

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-disj-biDFS.rkt}
	\end{tabular}
\end{center}

@@

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-conj-unfair.rkt}
	\end{tabular}
\end{center}

@@

BFS packages answers by their costs. The \emph{cost} of a answer is its 
depth in the search tree (i.e. the number of calls to relational definitions 
required to find them) \citet{seres1999algebra}. In the following example, 
every answer is a list of list of symbol. The cost of each of them is equal to 
the length of the inner lists plus the length of the outer list. In addition to 
being fair, BFS also produces answers in increasing order of cost.
\NOTE[MVC]{Here inner and outer are very confusing. Can you be more specified?}
\NOTE[LKC]{updated}

@@

BFS is fair in both \disj{} and \conj{}. Our implementation is based on 
DFS$_\textrm{f}$ (not DFS$_\textrm{i}$). All we have to do is apply two trivial changes to 
\texttt{append-map-inf}. First, rename it to \texttt{append-map-inf/fair}. 
Second, replace its use of \texttt{append-inf} to \texttt{append-inf/fair}.

>>>>

\newcommand{\conde}{\texttt{cond$^e$}}
\newcommand{\conj}{\texttt{conj}}
\newcommand{\disj}{\texttt{disj}}
\newcommand{\clisting}[1]{
\begin{center}
  \begin{tabular}{c}
	\lstinputlisting{#1}
    \end{tabular}
\end{center}
}
\newcommand{\DFSi}[0]{DFS$_\textrm{i}$}
\newcommand{\DFSf}[0]{DFS$_\textrm{f}$}
\newcommand{\BFS }[0]{BFS}

@@

The syntax of a programming language should reflect its semantics. When writing 
a \conde{} expression in miniKanren, a programmer would expect all clauses 
share the same chance of being explored, as these clauses are written in 
parallel. The existing search strategy, interleaving \DFSi{}, 
however, prioritize its clauses by the order how they are written down. 
Similarly, when a \conde{} is followed by another goal conjunctively, a 
programmer would expect answers in parallel share the same chance of being 
explored. Again, the answers by DFS$_\textrm{i}$ is different from the 
expectation. We have devised three new search strategies that have different 
level of fairness in \disj{} and \conj{}.

@@

% \conde is the flagship disjunction operator in miniKanren.
A subtlety arises when a \conde contains many clauses: not every clause has the 
equal chance to contribute to the result. As an example, consider the following 
relation \texttt{repeato} and its invocation.

\clisting{Figures/repeato.rkt}

Next, consider the following disjuction of invoking \texttt{repeato} with four 
different letters.

\clisting{Figures/example.rkt}

\conde intuitively corresponds to \texttt{or} in logic. And thus an unsuspicious
beginner would expect each letter to contribute equally to the result, as
following.

\clisting{Figures/run-repeato-fair.rkt}

The \conde in TR2, however, generates a less expected result.

\clisting{Figures/run-repeato-idfs.rkt}

The miniKanren in TR2 implements interleaving DFS(\DFSi), the cause of this 
unexpected result. With this search strategy, each \conde clause takes a half 
of its received computational resources and pass the other half to its 
following clauses, except for the last clause that takes all resources it 
receives. In the example above, the \texttt{a} clause takes half of all 
recourses. And the \texttt{b} clause takes a quarter. Thus \texttt{c} and 
\texttt{d} barely contribute to the result.

%In the example above, the letters \texttt{c} and \texttt{d} are 
%allocated with less resources than \texttt{a}, and thus \texttt{c} and 
%\texttt{d} barely contribute to the result.

\DFSi{} is sometimes powerful for an expert. By carefully organizing the order 
of \conde{} clauses, a miniKanren program can explore more ``interesting'' 
clauses than those uninteresting ones, and thus use computational resources 
efficiently. A little miniKanrener, however, may beg to differ--understanding
implementation details and fiddling with clauses order is not the first
priority of a beginner.

There is another reason that miniKanren could use more search strategies than
just \DFSi. In many applications, there does not exist one order that serves
for all purposes. For example, a relational dependent type checker contains
clauses for constructors that build data and clauses for eliminators that use
data. When this type checker is used to generate simple and shallow programs,
the clauses of constructors should be put in the front of \conde.
When performing proof search for complicated programs, the clauses of eliminator
should take the focus. With \DFSi, these two uses can not be efficient at the
same time. In fact, to make one use efficient, the other one must be drastically
slow.

The specification that every clause in the same \conde{} is given equal 
``search priority'' is called fair \disj{}. And search strategies with 
almost-fair \disj{} give every clauss in the same \conde{} similar priority. 
Fair \conj{}, a related concept, is more complicated. We defer it to the next 
section.

To summarize our contribution, we
\begin{itemize}
	\item analyze an existing concept, fair \disj{} by 
	\citet{seres1999algebra}.
	\item propose a new concept, almost-fair \disj{}.
	\item propose a new definition of fair \disj{}.
	\item propose and implement balanced interleaving depth-first search 
	(DFS$_\textrm{bi}$), a new search strategy with almost-fair \disj{}.
	\item propose and implement fair depth-first search (DFS$_\textrm{f}$), a 
	new search strategy with fair \disj{}.
	\item implement in a new way breath-first search (BFS), a search strategy 
	with fair \disj{} and fair \conj{}.
	(our code runs faster in all benchmarks and is simpler)
	\item prove our BFS implementation is equivalent with the one by 
	\citet{seres1999algebra}.
\end{itemize}

%\begin{figure}[tbp]
%	\begin{tabular}{|c|c|c|c|c|}
%		\hline 
%		fairness & DFS$_\textrm{i}$ & DFS$_\textrm{bi}$ & DFS$_\textrm{f}$ & 
%BFS \\ 
%		\hline 
%		disj & unfair & almost-fair & fair & fair \\ 
%		\hline 
%		conj & unfair & unfair & unfair & fair \\ 
%		\hline 
%	\end{tabular} 
%	\caption{fairness of search strategies}
%	\label{fairness}
%\end{figure}

@@

In this section, we define fair \disj{}, almost-fair \disj{} and fair \conj{}. 
Before going further into fairness, we would like to give a short review about 
state, search space, and goal, because fairness is defined in terms of them. 
Every answer given by miniKanren corresponds to a state. Each state records a 
way how some relations can be satisfied. In our case, the state is just a 
subsitution associating logic variables with terms. In more expressive 
miniKanrens (e.g. miniKanren with with symbolic constraints 
\citep{hemann2017framework}), states can include more information. A search 
space is a collection of states. And a goal is a function from a state to a 
search space. Goals check their input states and outputs possibly extended 
states. A goal might fail for some inputs, in which case the output search 
space would be empty.

Now we elaborate fairness by running more queries about \texttt{repeato}.

@@

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the answer. DFS$_\textrm{i}$, the current search
strategy, however, results in many more lists of \texttt{a}s than lists
of other letters. And some letters  (e.g. \texttt{c} and \texttt{d}) are
rarely seen. The situation would be exacerbated if \conde contains more clauses.

@@

Under the hood, the \conde here is allocating computational effort to 
four trivially different search space. The unfair \disj{} in 
\DFSi{} allocates much more effort to the first search space. On the contrary, 
fair \disj{} would allocate effort evenly to each search space.

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-disj-DFSf.rkt}
	\end{tabular}
\end{center}
\begin{center}
\begin{tabular}{c}
	\lstinputlisting{Figures/repeato-disj-BFS.rkt}
\end{tabular}
\end{center}

Running the same program again with almost-fair \disj 
(e.g. DFS$_\textrm{bi}$) gives the same result. Almost-fair, however, is not 
completely fair, as shown by the following example.

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-disj-DFSbi.rkt}
	\end{tabular}
\end{center}

DFS$_\textrm{bi}$ is fair only when the number of goals is a power of 2, 
otherwise, some goals would be allocated twice as many resources than the 
others. In the above example, where the \conde has five clauses, the clauses of 
\texttt{b}, \texttt{c}, and \texttt{d} are allocated more resources.

We end this subsection with precise definitions of all levels of 
\disj{} fairness. Our definition of \emph{fair} \disj{} is slightly generalize 
from the one by \citet{seres1999algebra}. Their definition is for binary 
disjunct. We generalize it to multi-arity one.

\begin{defn}[fair \disj{}]
	A \disj{} is fair iff it allocates computational resource evenly to 
	search spaces produced by goals in the same disjunct (e.g. clauses in the 
	same 
	\conde).
\end{defn}

\begin{defn}[almost-fair \disj{}]
	A \disj{} is almost-fair iff it allocates computational resource so 
	evenly to search spaces produced by goals in the same disjunct that the 
	maximal 
	ratio of resources is bounded by a constant.
\end{defn}

\begin{defn}[unfair \disj{}]
	A \disj{} is unfair iff it is not even almost-fair.
\end{defn}

@@

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-conj-DFSi.rkt}
	\end{tabular}
\end{center}

@@

BFS packages answers by their costs. The \emph{cost} of a answer is its 
depth in the search tree (i.e. the number of calls to relational definitions 
required to find them) \citet{seres1999algebra}. In the following example, 
every answer is a list of list of symbol. The cost of each of them is equal to 
the length of the inner lists plus the length of the outer list. In addition to 
being fair, BFS also produces answers in increasing order of cost.

@@

BFS has both fair \disj{} and fair \conj{}. Our implementation is based on 
DFS$_\textrm{f}$ (not DFS$_\textrm{i}$). To implement \BFS{} based on \DFSf{} 
all we have to do is renaming 
\texttt{append-map-inf} to \texttt{append-map-inf/fair}, then replacing its use 
of \texttt{append-inf} to \texttt{append-inf/fair}.
